Understanding Regional and Temporal Variations in Food Prices: Insights from US Consumer Price Index and Economic Trends

### 2024-11-26 paper/paper.qmd
updated analysis dataset, implemented eda and summary stats for outcome variable

We use the statistical programming language R [@citeR] to clean the data using various helpful packages like, readxl [@cite-readxl], dplyr [@cite-dplyr], tidyr [@cite-tidyr], lubridate [@cite-lubridate], and arrow [@cite-arrow].

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: out-summ-stats
#| tbl-cap: "Summary Statistics for log of Unit Value"

@@

The transformed outcome variable, unitValue_lg, represents the log-transformed weighted mean price per 100 grams of food items, with a mean of -0.775 and a median of -0.777, indicating a near-symmetric distribution. The standard deviation of 0.746 reflects moderate variability in prices, while the range, spanning from a minimum of -3.411 to a maximum of 1.191, highlights significant disparities in food costs across categories, regions, and time periods. The log transformation normalizes the data, reducing skewness and ensuring that extreme values in the original price variable do not disproportionately influence the analysis.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: out-summ-stats
#| tbl-cap: "Summary Statistics for log of Unit Value"

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: out-summ-stats
#| tbl-cap: "Summary Statistics for log of Unit Value"

@@

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

>>>>

We use the statistical programming language R [@citeR] to clean the data using various helpful packages like, readxl [@cite-readxl], dplyr [@cite-dplyr], tidyr [@cite-tidyr], lubridate [@cite-lubridate], and arrow [@cite-arrow]. Further, libraries like ggplot2 [@cite-gg], knitr [@cite-knitr] were used to analyze the data and create visualizations.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-outsumm
#| tbl-cap: "Summary Statistics for log of Unit Value"

@@

From @tbl-outsumm, we can see that the transformed outcome variable, unitValue_lg, has a mean of -0.692 and a median of -0.713, indicating a near-symmetric distribution. The standard deviation of 0.687 reflects moderate variability in prices, while the range, spanning from a minimum of -2.813 to a maximum of 1.191, highlights significant disparities in food costs across categories, regions, and time periods. The log transformation normalizes the data, reducing skewness and ensuring that extreme values in the original price variable do not disproportionately influence the analysis.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: hist-out
#| tbl-cap: "Distribution of Log-Transformed Food Prices (Unit Value)"

@@

@hist-out illustrates a symmetric, unimodal distribution of unitValue_lg, centered near -0.7, with values ranging from approximately -2.8 to 1.2. This indicates that most food prices cluster around the median value, with fewer extreme high or low prices.

More details and exploratory analysis about the outcome variable can be found in [Appendix -@sec-data-details]

@@

@@

## EDA for Outcome Variable

### Regional Trends

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: bxplt-out
#| tbl-cap: "Regional Variations in Log-Transformed Food Prices"

``` 
@bxplt-out reveals slight regional variations in log unit values. The Northeast and West regions show marginally higher median prices compared to the Midwest and South, reflecting higher food costs often associated with coastal or urban areas. The Midwest exhibits the largest variability, including a notable outlier on the high end, suggesting either an expensive food category or a regional anomaly. Despite these differences, the overall consistency in median prices across regions highlights a broadly stable pricing structure with minor geographic deviations. These trends emphasize the need to investigate the economic and geographic factors contributing to these regional disparities.

### Temporal Trends

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: plt-timetrend
#| tbl-cap: "Temporal Variations in Log-Transformed Food Prices"

transformed_data <- analysis_data %>% 
  mutate(time = time/12 + 2012) %>%
  group_by(time) %>%
  summarise(mean_unitValue_lg = mean(unitValue_lg, na.rm = TRUE))

ggplot(transformed_data, aes(x = time, y = mean_unitValue_lg)) +
  geom_line(color = "blue", size = 0.9) +
  labs(title = "Trend Over Time: Log-Transformed Unit Value of Food Prices",
       x = "Time (in Years)",
       y = "Log Unit Value") +
  theme_minimal()

@plt-timetrend illustrates the log-transformed unit value of food prices from 2012 to 2018, revealing a general upward trend over time. Initially, from 2012 to 2014, there is only a modest increase in the log-transformed values. However, in 2014, prices increase sharply followed by a period of 3 years characterized by heightened volatility, with pronounced peaks and troughs, and another increase in the prices in 2018. Thus, the overall trajectory for the log-transformed unit value of food prices from 2012 to 2018 remains upwards.

## EDA for Predictor Variables
